---
title: "Reproducible Research: Peer Assessment 1"
author: "MazinMelb"
date: "10 Jan 2016"
output: 
  html_document:
    keep_md: true
---
## Overview
This report is considering the activity data recorded by a device collecting data at 5 minute intervals through out the day. The data consists of two months activity from an anonymous individual collected over two months, October and November, 2012 ,and includes the number of steps taken in 5 minute intervals each day.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=5,warning=FALSE, message=FALSE)
```
```{r load_packages, include=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(stringr)
library(grid)
library(gridExtra)
```



## Loading and preprocessing the data

The data has been cleaned and transformed to facilitate visualisation and analysis. The time and date details have been converted into a 'r' date format (POSIXct) and there is a boolean field set to true for weekdays and false for weekends. 

```{r}
setwd("~/Documents/RepData_PeerAssessment1")
fname <- "activity.csv"
activities <- read.csv(fname)

# add boolean for weekend / weekday
activities <- activities %>%
              mutate(weekday=
                  substr(weekdays(as.POSIXct(date)),1,1)!="S")

# add POXIXct class version of the interval
timechar <- str_pad(as.character(activities$interval),4,side="left",pad="0") 
timeperiod <- sapply(timechar,function(x)
                    str_c(substr(x,1,2),":",substr(x,3,4)))

activities <- cbind(activities,timeperiod)
```


The data has been summarised by day, to show the total, mean and median number of steps by day. 
```{r}
actsummary <- activities %>%
              group_by(date) %>%
              summarise(totsteps=sum(steps,na.rm=TRUE),
                        msteps = mean(steps,na.rm=TRUE),
                        medsteps = median(steps,na.rm=TRUE),
                        weekday=unique(weekday))
```

It has also been summarised by 5-minute interval to show the daily pattern of activity.

```{r}
daysummary <- activities %>%
      group_by(timeperiod) %>%
      summarise(totsteps = sum(steps,na.rm=TRUE),
            msteps = mean(steps,na.rm=TRUE),
            medsteps=median(steps,na.rm=TRUE),
            interval=unique(interval))

maxmsteps <- filter(daysummary,msteps==max(msteps))
nbtext <- paste("  NB: Maximum (average \n  in a 5 minute interval) \n  is", 
                signif(maxmsteps$msteps,3),
                "steps at",maxmsteps$timeperiod)

# convert timeperiod into date/time format for graphing
maxmsteps$timeperiod=strptime(maxmsteps$timeperiod,format="%H:%M")
daysummary$timeperiod=strptime(daysummary$timeperiod,format="%H:%M")
``` 





# Summary of steps per day

## What is total and summarised number of steps taken per day (ignore NAs)?

Visualising the number of steps each day highlights the missing values. As can be seen in the graph below, the median number of steps per day is significantly affected by the missing data. The total number of steps, and the mean and median number of steps taken per day is shown below:

```{r, echo=FALSE}

# graph activities to show total steps per day
g1 <- ggplot(data.frame(actsummary), aes(x=as.Date(date),y=totsteps, fill=weekday))  + 
     geom_histogram(stat="identity",binwidth = 7) +
  xlab("Date") + ylab("Number of steps") +
  scale_x_date(labels = date_format("%d-%m-%Y"),breaks=date_breaks("week")) +
  theme(axis.text.x=element_text(angle=90)) +
  scale_fill_discrete(name = "Type of day",
                      breaks=c(FALSE,TRUE),
                      labels=c("Weekend","Weekday")) +
  ggtitle("Total number of steps (ignore NAs)") 

# graph activities to show total steps per day
g2 <- ggplot(data.frame(actsummary), aes(x=as.Date(date)))  + 
     geom_line(stat="identity",aes(y=medsteps, colour="Median steps")) +
     geom_line(stat="identity",aes(y=msteps, colour="Mean steps")) +
  xlab("Date") + ylab("Number of steps") +
  scale_x_date(labels = date_format("%d-%m-%Y"),breaks=date_breaks("week")) +
  scale_color_discrete(name="Summary of steps by interval")+
  theme(axis.text.x=element_text(angle=90)) +
  ggtitle("Mean, median of steps (ignore NAs)") 

  grid.arrange(g1,g2,ncol=2)
```


# Summary of daily activity pattern
## Daily pattern of activity (ignoring NAs)
The patern of daily activity, without imputing values for missing data, that is, any missing data has been ignormed, shows a difference in activity during the night hours (on the left and right of the graph), and the daytime hours (in the middle of the graph).

```{r, echo=FALSE}
# graph daily pattern of activity

g <- ggplot(daysummary,aes(x=timeperiod)) +
     geom_line(stat="identity",aes(y=medsteps, colour="Median steps")) +
     geom_line(stat="identity",aes(y=msteps, colour="Mean steps")) +
     annotate(geom="text", x=maxmsteps$timeperiod, y=maxmsteps$msteps,
              label=nbtext,hjust=0,vjust=1,size=3) +
     xlab("5-minute interval") + ylab("Number of steps") +
     theme(axis.text.x=element_text(angle=90)) +
     scale_x_datetime(labels=date_format("%H:%M"),        
                      breaks=date_breaks("2 hours"))  +
      scale_color_discrete(name="Summary of steps by interval")+
     ggtitle("Number of steps each interval (ignore NAs)") 
print(g)
```

## Imputing missing values

Examining the activity data suggests there are intervals and days where no activity data was logged. Assuming this is a gap in the data, rather than the highly unlikely complete and total absence of any trackable activity, the missing days and intervals will be replaced with the mean number of steps for that interval across all the logged days.

```{r}
# replace na with mean for that interval using a split apply combine approach
activitiesValue <- filter(activities,!is.na(steps))
activitiesNA <- filter(activities,is.na(steps))

# replace na values with mean for that interval
meansteps <- merge(activitiesNA,daysummary,by="interval") 
activitiesNA$steps <- meansteps$msteps

activities2 <- rbind(activitiesValue,activitiesNA)

# summarise steps per day with imputed values
actsummary2 <- activities2 %>%
               group_by(date) %>%
               summarise(totsteps=sum(steps,na.rm=TRUE),
                         msteps = mean(steps,na.rm=TRUE),
                         medsteps = median(steps,na.rm=TRUE),
                         weekday=unique(weekday))

# summarise steps per interval with imputed values
daysummary2 <- activities2 %>%
      group_by(timeperiod) %>%
      summarise(totsteps = sum(steps,na.rm=TRUE),
            msteps = mean(steps,na.rm=TRUE),
            medsteps=median(steps,na.rm=TRUE),
            interval=unique(interval))

daysummary2b <- activities2 %>%
                group_by(timeperiod, weekday) %>%
                summarise(totsteps = sum(steps,na.rm=TRUE),
                          interval=unique(interval))
daysummary2b$weekday <- sapply(daysummary2b$weekday, function(x)
                                ifelse(x,"Weekday","Weekend"))

maxmsteps <- filter(daysummary2,msteps==max(msteps))
nbtext <- paste("  NB: Maximum (average \n  in a 5 minute interval) \n  is", 
                signif(maxmsteps$msteps,3),
                "steps at",maxmsteps$timeperiod)

# convert timeperiod into date/time format for graphing
maxmsteps$timeperiod=strptime(maxmsteps$timeperiod,format="%H:%M")
daysummary2$timeperiod=strptime(daysummary2$timeperiod,format="%H:%M")
```


## Daily pattern of activity (imputed values)?

```{r, echo=FALSE}
# graph daily pattern of activity
g <- ggplot(daysummary2,aes(x=timeperiod,y=msteps,group=1)) +
     geom_line(stat="identity",aes(y=medsteps, colour="Median steps")) +
     geom_line(stat="identity",aes(y=msteps, colour="Mean steps")) +
     annotate(geom="text", x=maxmsteps$timeperiod, y=maxmsteps$msteps,
              label=nbtext,hjust=0,vjust=1,size=3) +
     xlab("5-minute interval") + ylab("Number of steps") +
     theme(axis.text.x=element_text(angle=90)) +
     scale_x_datetime(labels=date_format("%H:%M"),        
                      breaks=date_breaks("2 hours"))  +
      scale_color_discrete(name="Summary of steps by interval")+
     ggtitle("Number of steps each interval (imputed values)") 
print(g)
```

## What is total number of steps taken per day (imputed values)?

Visualising the number of steps each day after replacing NAs with the mean number of steps for that interval across days, shows a different pattern. As can be seen in the graph below, the median number of steps per day follows a pattern closer to the mean. 

The pattern of median number of steps per interval still varies from the mean number of steps per interval because of the number of 0 values recorded for some intervals. Without a better understanding of the recording mechanism it is unclear whether it would be appropriate to replace '0' values wiht the mean values.

The total number of steps, and the mean and median number of steps taken per day after replacing NAs with the mean values for that interval, are shown below:

```{r, echo=FALSE}

# graph activities to show total steps per day
g1 <- ggplot(data.frame(actsummary2), aes(x=as.Date(date),y=totsteps, fill=weekday))  + 
     geom_histogram(stat="identity",binwidth = 7) +
  xlab("Date") + ylab("Number of steps") +
  scale_x_date(labels = date_format("%d-%m-%Y"),breaks=date_breaks("week")) +
  theme(axis.text.x=element_text(angle=90)) +
  scale_fill_discrete(name = "Type of day",
                      breaks=c(FALSE,TRUE),
                      labels=c("Weekend","Weekday")) +
  ggtitle("Total number of steps (imputed values)") 

# graph activities to show total steps per day
g2 <- ggplot(data.frame(actsummary2), aes(x=as.Date(date)))  + 
     geom_line(stat="identity",aes(y=medsteps, colour="Median steps")) +
     geom_line(stat="identity",aes(y=msteps, colour="Mean steps")) +
  xlab("Date") + ylab("Number of steps") +
  scale_x_date(labels = date_format("%d-%m-%Y"),breaks=date_breaks("week")) +
  scale_color_discrete(name="Summary of steps by interval")+
  theme(axis.text.x=element_text(angle=90)) +
  ggtitle("Mean, median of steps (imputed values)") 

  grid.arrange(g1,g2,ncol=2)
```



## Are there differences in activity patterns between weekdays and weekends?

The graph below visualises the different activities patterns between weekdays and weekends. As might be expected, there is less activty on the weekends and more activity (measured in total number of steps for each interval) on weekdays.

```{r, echo=FALSE}
# graph daily pattern of activity
g <- ggplot(daysummary2b,aes(x=interval,y=totsteps)) +
     geom_line(stat="identity") +
     xlab("5-minute interval") + ylab("Number of steps") +
     theme(axis.text.x=element_text(angle=90)) +
     ggtitle("Number of steps each interval (imputed values)") +
     facet_grid(weekday ~ .)
print(g)
```

